---
title: "An example for use of goSorensen R-package"
author: ""
package: goSorensen
abstract: >
 This document provides an example for the manipulation of `goSorensen` R Package, using  feature lists  achieved from real a real microarray experiment.
 
  Starting from a fast introduction about the installation of package and description of the data, we explain how to: i) perform the equivalence test from contingency tables of joint enrichment or directly from features lists (either using a normal asymptotic or a bootstrap approximation), ii) collect specific fields of the test results like the p-value, the upper limit of the confidence interval or standard errors  iii) obtain another statistics related to the Sorensen-Dice dissimilarity, iv) compute new results for updated input values (i.e. confidence level, irrelevance limit, used distribution, etc) and v) perform all pairwise computes
  
  This document is based on the vignette "_An introduction to package goSorensen_" available on the `goSorensen` R-Package
  
output:
  BiocStyle::html_document:
    toc_float: true
vignette: >
  %\VignetteIndexEntry{An introduction to equivalence test between feature lists using goSorensen.}
  %\VignetteEngine{knitr::rmarkdown}
  %%\VignetteKeywords{Annotation, GO, GeneSetEnrichment, Software, Microarray, Pathways, GeneExpression, MultipleComparison}
  %\VignetteEncoding{UTF-8}
---

```{r style, echo = FALSE, results = 'asis'}
BiocStyle::markdown()
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(dpi=25,fig.width=7)
```

```{r env, message = FALSE, warning = FALSE, echo = TRUE}
library(goSorensen)
```


# Installation

goSorensen package has to be installed with a working R version `(R >= 4.2.0)`. Installation could take a few minutes on a regular desktop or laptop. Package can be installed from Bioconductor or `devtools` package, then it needs to be loaded using `library(goSorensen)`

To install from Bioconductor (recommended):

```{r, eval=FALSE}
## Only if BiocManager is not previosly installed:
install.packages("BiocManager")

## otherwise, directly:
BiocManager::install("goSorensen")
```

To install from Github

```{r, eval=FALSE}
devtools::install_github("pablof1988/goSorensen", build_vignettes = TRUE)
```

# Data.
`allOncoGeneLists` is a comprehensive set of gene lists related to cancer obtained from <http://www.bushmanlab.org/links/genelists>. `goSorensen` package save in a "list" object the seven largest lists and these can be loaded by means of  `data("allOncoGeneLists")`
```{r}
data("allOncoGeneLists")
```


Names and length of the gene lists to be compared are:
```{r}
sapply(allOncoGeneLists, length)
```

# Performing the equivalence test

## Building 2 x 2 contingency table of mutual enrichment
It is possible to build the enrichment contingency table between gene lists (for example "Sanger" and "Vogelstein") for an specific Ontology and GO level (for example Ontology BP and GO level 4)

```{r, message=FALSE, warning=FALSE}
library(goSorensen)
data("humanEntrezIDs") # necessary to recognize features in ENTREZ id's
enrichTab <- buildEnrichTable(allOncoGeneLists$sanger,
                              allOncoGeneLists$Vogelstein,
                              geneUniverse = humanEntrezIDs, 
                              orgPackg = "org.Hs.eg.db",
                              onto = "BP", GOLevel = 4, 
                              listNames = c("Sanger", "Vogelstein"))
enrichTab
```

## Equivalence test

### Equivalence test from contingency table
Equivalence test from contingency table for an equivalence (or irrelevance) limit $d_0=0.2857$ and a significance level $\alpha = 0.05$
```{r}
contin_testResult <- equivTestSorensen(enrichTab, d0 = 0.2857, 
                                       conf.level = 0.95)
contin_testResult 
```
This process is using by default the normal approximation to the sample distribution of the $(\widehat{d}−d)/\widehat{se}$ statistic. Alternatively, it is possible to estimate this distribution by means of bootstrap:
```{r}
boot_testResult <- equivTestSorensen(enrichTab, d0 = 0.2857, 
                                       conf.level = 0.95, boot = T)
boot_testResult 
```
 For low frequencies in the contingency table, bootstrap is a more conservative  but preferable approach, with better type I error control.

### Equivalence test from feature lists.
Obtain the test directly from gene lists is also possible:
```{r}
lists_testResult <- equivTestSorensen(allOncoGeneLists$sanger,
                                      allOncoGeneLists$Vogelstein, 
                                      d0 = 0.2857,
                                      geneUniverse = humanEntrezIDs, 
                                      orgPackg = "org.Hs.eg.db",
                                      onto = "BP", GOLevel = 4, 
                                      listNames = c("Sanger", "Vogelstein"))
lists_testResult
```

# Accessing to specific fields
To access specific fields from the test result:
```{r}
# Contingency table from equivalence test using normal approximation:
getTable(lists_testResult) 

# Sorensen-Dice dissimilarity from equivalence test using normal approximation:
getDissimilarity(lists_testResult) 

# p value from equivalence test using normal approximation:
getPvalue(lists_testResult) 

# p value from equivalence test using bootstrap approximation:
getPvalue(boot_testResult) 
```

and the same for other specific fields: standard error `getSE`, upper bound equivalence limit `getUpper` and efective number of bootstrap and resamples `getNboot` (Only available for bootstrap tests)

# Other statistics related to the Sorensen-Dice dissimilarity
Sometimes, it would be interesting not to perform the full equivalence test  but to compute other statistics related to the Sorensen-Dice dissimilarity. This computes would be done from contingency tables or directly from feature lists:

- The dissimilarity:
```{r}
dSorensen(enrichTab)
dSorensen(allOncoGeneLists$sanger,
          allOncoGeneLists$Vogelstein,
          geneUniverse = humanEntrezIDs, 
          orgPackg = "org.Hs.eg.db",
          onto = "BP", GOLevel = 4, 
          listNames = c("Sanger", "Vogelstein"))

```

- The Upper limit of the confidence interval for the true distance:
```{r}
duppSorensen(enrichTab)
duppSorensen(enrichTab, boot = T)
duppSorensen(allOncoGeneLists$sanger,
             allOncoGeneLists$Vogelstein,  
             geneUniverse = humanEntrezIDs, 
             orgPackg = "org.Hs.eg.db",
             onto = "BP", GOLevel = 4, 
            listNames = c("Sanger", "Vogelstein"))
```

- the same for `seSorensen`

# Updating the results:
When some inputs (i.e. confidence level, irrelevance limit, used distribution, etc) have to be updated, it is no necessary to make computes again, only updating is enough
```{r}
upgrade(lists_testResult, d0 = 0.175, conf.level = 0.99, boot = T)
```

# All pairwise tests (or other computations)
For objects of class list, all these functions (`equivTestSorensen`, `dSorensen`,  `seSorensen`, `duppSorensen`) assume a list of character objects containing gene identifiers and all pairwise computations are performed. 

For example, to obtain the matrix of all pairwise Sorensen-Dice dissimilarities:
```{r}
dSorensen(allOncoGeneLists, onto = "BP", GOLevel = 4, 
          geneUniverse = humanEntrezIDs, orgPackg = "org.Hs.eg.db")
```

Similarly, the following code performs all pairwise tests for all Ontologies and GO levels. In this case the Holm-Bonferroni criteria to avoid the inflation of p values for multiple tests is used:
```{r, eval=FALSE}
allOncoResults <- equivTestSorensen(allOncoGeneLists, 
                                    geneUniverse = humanEntrezIDs,
                                    orgPackg = "org.Hs.eg.db")
```
Visualization of  these computes is not good due to the huge number of results obtained, for this reason they were saved in the available file "allOncoResults.rda" using  `save(allOncoResults, file = "allOncoResults.rda")`. If user wishes to see these results, he can easily load them using `load("allOncoResults.rda")`. In addition, files: "summary_goSorensen.xlsx", "AllResults_Test.txt" and "AllResults_Tables.txt" are also provided with a summary of these results.

Remember that, it is possible to access to specific fields of interest for specifics Ontologies and GO levels, for example:
```{r}
load("allOncoResults.rda")
allOncoResults$BP$`level 4`$Vogelstein$sanger
getPvalue(allOncoResults$BP$`level 4`$Vogelstein$sanger)
```

# Session information {.unnumbered}

All software and respective versions used to produce this document are listed below.

```{r sessionInfo}
sessionInfo()
```

# References {.unnumbered}

Flores, P., Salicrú, M., Sánchez-Pla, A. et al. An equivalence test between features lists, based on the Sorensen--Dice index and the joint frequencies of GO term enrichment. BMC Bioinformatics 23, 207 (2022). <https://doi.org/10.1186/s12859-022-04739-2>
