---
title: "An example for use of goSorensen R-package"
author: ""
package: goSorensen
abstract: >
 This document provides an example for the manipulation of `goSorensen` R Package, using  feature lists  achieved from a real microarray experiment.
 
  Starting from a fast introduction about the installation of package and description of the data, we explain how to: i) perform the equivalence test from contingency tables of joint enrichment or directly from feature lists (either using a normal asymptotic or a bootstrap approximation to the sampling distribution), ii) collect specific fields of the test results like the p-value, the upper limit of the confidence interval for the Sorensen-Dice dissimilarity or the standard errors of the dissimilarity estimates  iii) obtain additional statistics related to the Sorensen-Dice dissimilarity and iv) update preexisting test results (e.g., how to change the confidence level, the limit of equivalence or the testing method -normal approximation or bootstrap)
  
  This document is based on the vignette "_An introduction to package goSorensen_" available on the `goSorensen` R-Package
  
output:
  BiocStyle::html_document:
    toc_float: true
vignette: >
  %\VignetteIndexEntry{An introduction to equivalence test between feature lists using goSorensen.}
  %\VignetteEngine{knitr::rmarkdown}
  %%\VignetteKeywords{Annotation, GO, GeneSetEnrichment, Software, Microarray, Pathways, GeneExpression, MultipleComparison}
  %\VignetteEncoding{UTF-8}
---

```{r style, echo = FALSE, results = 'asis'}
BiocStyle::markdown()
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(dpi=25,fig.width=7)
```

```{r env, message = FALSE, warning = FALSE, echo = TRUE}
library(goSorensen)
```


# Installation

The goSorensen package has to be installed in a working R version `(R >= 4.2.0)`. Installation could take a few minutes on a regular desktop or laptop. The package can be installed from Bioconductor or from GitHub.

To install it from Bioconductor (recommended):

```{r, eval=FALSE}
## Only if BiocManager is not previosly installed:
install.packages("BiocManager")

## otherwise, directly:
BiocManager::install("goSorensen")
```

To install from GitHub:
```{r, eval=FALSE}
devtools::install_github("pablof1988/goSorensen", build_vignettes = TRUE)
```

Finally, load the package to use it:
```{r}
library(goSorensen)
```

# Data.
`allOncoGeneLists` is a collection of seven gene lists related to cancer obtained from <http://www.bushmanlab.org/links/genelists>. They are included in the package and can be loaded by means of
```{r}
data("allOncoGeneLists")
```

Names and length of the 7 gene lists:
```{r}
sapply(allOncoGeneLists, length)
```

# Performing the equivalence tests

## Building the 2 x 2 contingency table of mutual enrichment:
It is possible to build the enrichment contingency table between gene lists (for example "Sanger" and "Vogelstein") for an specific Ontology and GO level (for example Ontology BP and GO level 4)

```{r, message=FALSE, warning=FALSE}
data("humanEntrezIDs") # necessary to recognize  ENTREZhuman gene identifiers
enrichTab <- buildEnrichTable(allOncoGeneLists$sanger,
                              allOncoGeneLists$Vogelstein,
                              geneUniverse = humanEntrezIDs, 
                              orgPackg = "org.Hs.eg.db",
                              onto = "BP", GOLevel = 4, 
                              listNames = c("Sanger", "Vogelstein"))
enrichTab
```

But it is also possible to build all pairwise enrichment contingency tables for all the gene lists of data set (`allOncoGeneLists` in this case) for an specific Ontology and GO level (for example Ontology BP and GO level 4):
```{r, eval=FALSE}
# The next code spends a lot of time due to the huge quantity of computed 
# results. Instead of running this line, we suggest loading the results which 
# are available in the file "allEnrichTable.rda" using:
# load("allEnrichTable.rda")
allEnrichTable <- buildEnrichTable(allOncoGeneLists, 
                                   geneUniverse = humanEntrezIDs, 
                                   orgPackg = "org.Hs.eg.db",
                                   onto = "BP", GOLevel = 4)
```

Of course, it is possible to filter an specific contingency table:
```{r}
load("allEnrichTable.rda")
allEnrichTable$miscellaneous$atlas
```

## Equivalence test

### Equivalence test from the contingency table
Equivalence test from this contingency table for an equivalence (or irrelevance) limit $d_0=0.2857$ and a significance level $\alpha = 0.05$
```{r}
contin_testResult <- equivTestSorensen(enrichTab, d0 = 0.2857, 
                                       conf.level = 0.95)
contin_testResult 
```
By default, this function uses a normal approximation for the true sampling distribution of the test statistic $(\widehat{d}−d)/\widehat{se}$. Alternatively, it is possible to estimate this distribution by means of bootstrap:
```{r}
boot_testResult <- equivTestSorensen(enrichTab, d0 = 0.2857, 
                                     conf.level = 0.95, boot = TRUE)
boot_testResult 
```
 For low frequencies in the contingency table, the bootstrap is a more conservative  but preferable approach, with better type I error control.

### Equivalence test directly from two gene lists:
To obtain the test directly from gene lists is also possible:
```{r}
lists_testResult <- equivTestSorensen(allOncoGeneLists$sanger,
                                      allOncoGeneLists$Vogelstein, 
                                      d0 = 0.2857,
                                      geneUniverse = humanEntrezIDs, 
                                      orgPackg = "org.Hs.eg.db",
                                      onto = "BP", GOLevel = 4, 
                                      listNames = c("Sanger", "Vogelstein"))
lists_testResult
```

# Accessing specific fields of the result
To access specific fields from the test result:
```{r}
# Contingency table:
getTable(lists_testResult) 

# Sorensen-Dice dissimilarity:
getDissimilarity(lists_testResult) 

# p-value from equivalence test using normal approximation:
getPvalue(lists_testResult) 

# p value from equivalence test using bootstrap approximation:
getPvalue(boot_testResult) 
```

and the same for other specific fields: standard error `getSE`, upper bound equivalence limit `getUpper` and efective number of bootstrap and resamples `getNboot` (only available for bootstrap tests)

# Other statistics related to the Sorensen-Dice dissimilarity
Sometimes, it would be interesting not to perform the full equivalence test  but to compute other statistics related to the Sorensen-Dice dissimilarity. These computations would be done from contingency tables or directly from feature lists:

- The dissimilarity:
```{r}
dSorensen(enrichTab)
dSorensen(allOncoGeneLists$sanger,
          allOncoGeneLists$Vogelstein,
          geneUniverse = humanEntrezIDs, 
          orgPackg = "org.Hs.eg.db",
          onto = "BP", GOLevel = 4, 
          listNames = c("Sanger", "Vogelstein"))

```

- The Upper limit of the confidence interval for the true distance:
```{r}
duppSorensen(enrichTab)
duppSorensen(enrichTab, boot = T)
duppSorensen(allOncoGeneLists$sanger,
             allOncoGeneLists$Vogelstein,  
             geneUniverse = humanEntrezIDs, 
             orgPackg = "org.Hs.eg.db",
             onto = "BP", GOLevel = 4, 
             listNames = c("Sanger", "Vogelstein"))
```

- the same for `seSorensen`

# Updating the results:
When some inputs (i.e. confidence level, irrelevance limit, used sampling distribution -normal or bootstrap-, etc) have to be updated, it is no necessary to recompute again from scratch, just update:
```{r}
upgrade(lists_testResult, d0 = 0.175, conf.level = 0.99, boot = TRUE)
```

# All pairwise tests (or other computations)
For objects of class "list", all these functions (`equivTestSorensen`, `dSorensen`,  `seSorensen`, `duppSorensen`) assume a list of character objects containing gene identifiers and all pairwise computations are performed. 

For example, to obtain the matrix of all pairwise Sorensen-Dice dissimilarities:
```{r}
dSorensen(allOncoGeneLists, onto = "BP", GOLevel = 4, 
          geneUniverse = humanEntrezIDs, orgPackg = "org.Hs.eg.db")
```

Similarly, the following code performs all pairwise tests for the  ontologies ( "BP", "MF" and "CC", all by default) and GO levels (from 3 to 10 by default) desired

```{r, eval=FALSE}
# The next code spends a lot of time due to the huge quantity of computed 
# results. Instead of running this line, we suggest loading the results which 
# are available in the file "allOncoResults.rda" using  
# load("allOncoResults.rda")
allOncoResults <- allEquivTestSorensen(allOncoGeneLists, 
                                    geneUniverse = humanEntrezIDs,
                                    orgPackg = "org.Hs.eg.db")
```
In addition, files: "summary_goSorensen.xlsx", "AllResults_Test.txt" and "AllResults_Tables.txt" are also provided with a summary of these results.

Remember that, it is possible to access to specific fields of interest for specifics Ontologies and GO levels, for example:
```{r}
load("allOncoResults.rda")

# Results for an specific ontology and level:
allOncoResults$BP$`level 4`$Vogelstein$sanger

# p value for an specific ontology and level:
getPvalue(allOncoResults, onto = "BP", GOLevel = 4, 
          listNames = c("Vogelstein","sanger"))

# Display all p-values: 
# getPvalue(allOncoResults)

# Display all upper confidence limits for the Sorensen-Dice dissimilarity
# getUpper(allOncoResults)
```


# Session information {.unnumbered}

All software and respective versions used to produce this document are listed below.

```{r sessionInfo}
sessionInfo()
```

# References {.unnumbered}

Flores, P., Salicrú, M., Sánchez-Pla, A. et al. An equivalence test between features lists, based on the Sorensen--Dice index and the joint frequencies of GO term enrichment. BMC Bioinformatics 23, 207 (2022). <https://doi.org/10.1186/s12859-022-04739-2>
